---
title: "Rxnat - Download full subject data"
author: "Adi Gherman"
date: "9/17/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Download a full set of data for a specific subject from the NITRC XNAT repository

The following tutorial will provide a fully functional example on how to download a full set of images for a specific subject, using the Rxnat package functionality.

### Install the Rxnat package
You can install `Rxnat` from Neuroconductor with:
``` {r, eval = FALSE}
source("https://neuroconductor.org/neurocLite.R")
neuro_install('Rxnat')
```

### Setting up system environment variables
To use system environment variables we need to add them to the `.Renviron` file located in the user's home folder. Use `Sys.getenv("HOME")` to get the path (for unix/osx users the location can be easily accessed with ~, eg. `vi ~/.Renviron`).
The `Rxnat` package will be able to automatically read / use a system environment variable provided the following format is used: `XXXX_RXNAT_USER` and `XXXX_RXNAT_PASS`. `XXXX` is provided as an argument when an XNAT connection is initiated. 

As an example `NITRC` is used as argument and the system environment variables names should be `NITRC_RXNAT_USER`, and `NITRC_RXNAT_PASS`.
``` {r, eval = TRUE}
library(Rxnat)
library(dplyr)
nitrc <- xnat_connect('https://nitrc.org/ir', xnat_name='NITRC')
```

### Retrieve list of subjects
To retrieve the full list of subjects for the `nitrc` XNAT connection, we will use the `nitrc$subjects()` function.
``` {r, eval = TRUE}
nitrc_subjects <- nitrc$subjects()
head(nitrc_subjects)
```

### Retrieve list of experiments
To retrieve the full list of experiments for the `nitrc` XNAT connection, we will use the `nitrc$experiments()` function.
``` {r, eval = TRUE}
nitrc_experiments <- nitrc$experiments()
head(nitrc_experiments)
```

### Get list of all experiments associated with a subject
For the purpose of this example, we would like to identify a subject that has more than one experiments associated with his id. In order to figure out a good subject ID example, we will filter and select all subjects with more than 1 experiment associated.
``` {r, eval = TRUE}
multi_session = nitrc_experiments %>% 
  group_by(subject) %>% 
  mutate(n_sessions = n()) %>% 
  filter(n_sessions > 1)
head(multi_session)
```

We will select the first subject ID (`UCLA_51209`) to continue our example. 
``` {r, eval = TRUE}
subject_id <- "UCLA_51209"
subject_experiments = nitrc_experiments[ nitrc_experiments$subject %in% subject_id, ]
head(subject_experiments)
```

<!--
### Retrieve all resources associated with a subject ID
Now that we have all the experiments associated with a subject ID, we can list all the available resources. 
``` {r, eval=TRUE}
all_resources = lapply(subject_experiments$ID, function(r) {
  nitrc$get_xnat_experiment_resources(experiment_ID = r)
})
head(all_resources)
```
-->

### Download all resources
We will use the `download_dir()` function to download of all resources associated with the selected subject ID.
``` {r, eval = TRUE}
download_data = lapply(subject_experiments$ID, function(r) {
  nitrc$download_dir(experiment_ID = r, extract = TRUE)
})
head(download_data)
```